The role of Similarity in native spoken word recognition

### 2024-10-08 manuscript/manuscript.qmd
Add discussion and data about knowledge filter

# import objects
tar_config_set(store = here::here("_targets"),
               script = here::here("_targets.R"))
tar_load_globals()
tar_load(participants)
tar_load(exp_participants)
tar_load(exp_responses)
tar_load(quest_responses)
tar_load(quest_participants)
tar_load(stimuli)
tar_load(epreds)
tar_load(dataset_1)
tar_load(dataset_2)
tar_load(dataset_3)
tar_load(exp_1_m0)
tar_load(exp_2_m0)
tar_load(exp_3_m1)

theme_set(theme_ggdist() + 
              theme(panel.grid.major.y = element_line(colour = "grey",
                                                      linetype = "dotted")))

@@

When some German speakers listened to the song *The Power* by SNAP! (*Coyote Ugly*, 2000), many of them misheard the line "I've got the power" as *Agatha Bauer!* When Michael Jackson vocalised "ma-ma-coo-sah" in his song *Wanna Be Startin' Somethin'* (*Thriller*, 1982), some Dutch speakers understood *mama appelsap* [mama applejuice]. When Spanish speakers listen to the line "Circumvent your thick ego" from the song *Pictures* by System of a Down (*Steal this Album!*, 2002), they tend to hear *Sácame de aquí, King-Kong* [take me out of here, King-Kong]. When Japanese speakers listen to the line "I want to hold your hand" in the homonym song by The Beatles (*Meet The Beatles*, 1964), they often mishear *Aho na hōnyōhan* [stupid urinator]. Outrageous as these examples might sound to speakers of other languages, many readers may know a few cases in their own native language. This auditory illusion can feel quite real, and often inevitable after the first time it happens. In Japanese, this phenomenon takes the name *Soramimi* (lit. "empty ear"), after the *Soramimi Hour*---a section of the program *Tamori Club* hosted by comedian Morita Kazuyoshi (Tamori)---in which instances of this illusion were presented with comedy purposes. Soramimi is a particular case of *homophonic translation*: words or phrases in one language are translated into similar-sounding words and phrases in a different language, without necessarily preserving the meaning. This phenomenon extends beyond song lyrics, and may occur anytime when listening to non-native speech. For instance, when listening to "You know me?" (\textipa{/ju: n@U mi:}/) in English, Japanese-native listeners tend to understand *yunomi* (/\textipa{jUnomi}/), which is a type of Japanese teacup. Overall, the occurence of Soramimi---and more generally, homophonic translations---suggests that speech in an unfamiliar language has the potential to activate lexical representations in the native language.

Although homophonic translations do not necessarily take place with the preservation of meaning---which often results in outrageous mistranslations like the ones aforementioned---homophonic translations sometimes may lead to the activation of the intended meaning (i.e., a correct translation). For example, imagine a Spanish native with no previous familiarity with any other language listening to French for the first time. This listener may encounter the word /\textipa{"pOKt}/ (*porte*), translation of door in French. When comparing the phonemic transcription of the French to that of its Spanish translation /\textipa{"pweR.ta}/ (*puerta*), one may be led to infer that these two words do not have much overlap. But one thing to note is that the voiced uvular fricative consonant /\textipa{K}/ and the open-mid back rounded vowel /\textipa{O}/, which do not exist in Spanish as phonemic categories, can be perceived as allophonic variations the native (Spanish) phonemes /\textipa{R}/ and /o/, as described by assimilation models of phoneme perception [@best2001discrimination; @best1988examination]. Although this phonetic mismatch between the non-native word and its translation can have a noticeable toll on comprehension @cutler2004patterns, the Spanish listener may still be able to activate the phonological representation of /\textipa{"pweR.ta}/ (*puerta*) from hearing the word-form /\textipa{"pOKt}/ (*porte*), ultimately allowing them to access the correct semantic representation. Given enough phonological similarity between the non-native word and its translation in the native language, a naïve listener may succeed at word comprehension when listening to an unfamiliar language. Such phonological similarity between translation equivalents---known as *cognateness*---is common across many languages, and is often due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents  [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and orthographic similarity of the former to the equivalent in their native language.

While Soramimi, and more generally homophonic translation, are popular and frequent phenomena, the psycholinguistic literature on the topic is relatively scarce, and has mostly focused on the phonological variables involved in the occurence of Soramimi. For instance, @otake2007interlingual analysed 194 instances of Soramimi, broadcasted between 1992 and 2007 by the TV show Soramimi hour. These instances consisted of homophonic translations of English song lyrics to Japanese words and phrases. The vast majority of the reported instances were phrases (96%) and a few, to single words (4%). The phonetic features of the presented English lyrics were preserved with relatively high variability, with some Japanese resulting words or phrases sharing high overlap with the original English strings of sounds, and some sharing very little overlap. When analysing the few instances of homophonic translations of single words, the author identified three phonological processes that might explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /\textipa{kraI}/ to *kurai*, /\textipa{kuRai}/ [dark]), deletion (e.g., *go*, /\textipa{goU}/ to *go*, /go/ [go], and alternation (e.g., *low*, /\textipa{loU}/ to *rou*, /\textipa{Roo}/ [wax]). This suggests that the reported homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to the Japanese phoneme inventory and Japanese phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic]. Native prosody also constrains the occurence of homophonic translations. For instance, @kentner2015rhythmic presented in the auditory modality French-native and German-native participants with English songs with lyrics. Homophonic translations followed language-specific segmentation strategies. Most homophonic translations reported by German-native participants resulted from inserting phrase and word boundaries *before* stressed syllables (following the more frequent trochaic patter in German). On the contrary, most homophonic translations reported by French-native participants resulted from inserting phrase and word boundaries *after* stressed syllables, following the more frequent iambic pattern in French.

@@

@@

```{r}
#| label: participants-numbers-1
participants_1 <- exp_participants |> 
    filter(group!="cat-SPA")

@@

n_participants <- n_distinct(participants_1)
n_participants_group <- table(participants_1$group)
participants_age <- participants_1 |> 
    summarise(across(age, lst(mean, sd, min, max))) 
n_female <- table(participants_1$gender)
```

Participants were recruited via Prolific (5£ compensation) and SONA (compensation in academic credits), and gave informed consent before providing any data and the study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was by the University of Oxford Medical Sciences Inter-Divisional Research Ethics Committee (IDREC) (R60939/RE007). Participants were asked to complete the experiment using a laptop in a quiet place with good internet connection.  Data collection took place from `r collection_dates[1]` to `r collection_dates[2]`. We collected data from `r n_participants` British English native participants living in United Kingdom (*Mean* = `r round(participants_age$age_mean, 2)` years, *SD* = `r round(participants_age$age_sd, 2)`, Range = `r round(participants_age$age_min, 2)`-`r round(participants_age$age_max, 2)`, `r n_female` female).

```{r tbl-participants}
#| label: tbl-participants
tbl_data <- participants |> 
    mutate(exp = case_when(
        source=="Experiment" & group!= "cat-SPA" ~ "Experiment 1",
        source=="Experiment" & group== "cat-SPA" ~ "Experiment 2",
        source=="Questionnaire" ~ "Experiment 3",
    )) |> 
    summarise(n = n(),
              n_excluded = sum(!valid_status=="Valid"),
              across(age, lst(mean, sd, min, max)),
              l2_lst = list(l2),
              .by = c(group, exp)) |> 
    mutate(l2_lst = map(l2_lst, 
                        function(x) {
                            y <- table(x)
                            y <- y[names(y) != "None"]
                            y <- paste0(names(y), " (", y, ")")
                            return(y)
                        }),
           n_excluded = paste0(n, " (", n_excluded, ")")) |> 
    select(-n)

tbl_data |> 
    gt(groupname_col = "exp", rowname_col = "group") |> 
    fmt_number(matches("age")) |> 
    fmt_integer(matches("min|max")) |> 
    cols_merge_uncert(age_mean, age_sd) |> 
    cols_merge_range(age_min, age_max) |> 
    tab_spanner("Age", matches("age")) |> 
    cols_label(group = "Group",
               exp = "Experiment",
               n_excluded = "N",
               age_mean = "Mean ± SD ",
               age_min = "Range",
               l2_lst = "L2") |> 
    tab_style(cell_text(style = "italic"),
              cells_column_labels()) |> 
    tab_style(cell_text(weight = "bold"),
              cells_column_spanners()) |> 
    tab_style(cell_text(align = "left"),
              list(cells_body(), cells_column_labels())) |> 
    tab_style(cell_text(weight = "bold",
                        align = "center",
                        style = "normal"),
              cells_column_labels(c(l2_lst, n_excluded))) |> 
    tab_footnote(
        "Number of included participants (number of excluded participants.)", 
        locations = cells_column_labels(n_excluded))
```

@@

```{r}
#| label: stimuli-lengths
lengths <- stimuli |>  
    select(group, ipa_flat_1, ipa_flat_2, word_1, word_2) |>  
    mutate(across(ipa_flat_1:word_2, nchar)) |> 
    summarise(across(c(ipa_flat_1, ipa_flat_2, word_1, word_2), 
                     lst(mean, sd, min, max)),
              .by = group)
lengths <- split(lengths, lengths$group)
names(lengths) <- janitor::make_clean_names(names(lengths))

durations <- stimuli |> 
    summarise(across(duration, lst(mean, sd, min, max)),
              .by = group) 
durations <- split(durations, durations$group)
names(durations) <- janitor::make_clean_names(names(durations))
```

We arranged two lists of input words to be presented to participants in the auditory modality: one in Catalan and one in Spanish. Words in the Catalan list were `r round(lengths$cat_eng$ipa_flat_1_mean, 2)` phonemes long on average (*SD* = `r round(lengths$cat_eng$ipa_flat_1_sd, 2)`, Range = `r round(lengths$cat_eng$ipa_flat_1_min, 2)`-`r round(lengths$cat_eng$ipa_flat_1_max, 2)`), and the orthographic forms of their English translations (which participants had to type) were `r round(lengths$cat_eng$word_2_mean, 2)` characters long on average (*SD* = `r round(lengths$cat_eng$word_2_sd, 2)`, Range = `r round(lengths$cat_eng$word_2_min, 2)`-`r round(lengths$cat_eng$word_2_max, 2)`. Words in the Spanish list were `r round(lengths$spa_eng$ipa_flat_1_mean, 2)` phonemes long on average (*SD* = `r round(lengths$spa_eng$ipa_flat_1_sd, 2)`, Range = `r round(lengths$spa_eng$ipa_flat_1_min, 2)`-`r round(lengths$spa_eng$ipa_flat_1_max, 2)`), and the orthographic form of their English translations (which participants had to type), were `r round(lengths$spa_eng$word_2_mean, 2)` characters long on average (*SD* = `r round(lengths$spa_eng$word_2_sd, 2)`, Range = `r round(lengths$spa_eng$word_2_min, 2)`-`r round(lengths$spa_eng$word_2_max, 2)`.

Participants listened to one audio file in each trial, each containing a single word presented in isolation. The audio files were the same ones used in child experiments conducted in the Laboratori de Recerca en Infància of Universitat Pompeu Fabra (Barcelona, Spain). These audio files were recorded by a proficient Catalan-Spanish female bilingual from the Metropolitan Area of Barcelona in a child-directed manner. Catalan and Spanish words were recorded at 44,100 Hz in separate files in the same session, and then de-noised using Audacity and normalised at peak intensity using Praat [@broersma2021praat]. The average duration of the Catalan audio files was `r round(durations$cat_eng$duration_mean, 2)` (*SD* = `r round(durations$cat_eng$duration_sd, 2)`, Range = `r round(durations$cat_eng$duration_min, 2)`-`r round(durations$cat_eng$duration_max, 2)`). The average duration of the Catalan audio files was `r round(durations$spa_eng$duration_mean, 2)` (*SD* = `r round(durations$spa_eng$duration_sd, 2)`, Range = `r round(durations$spa_eng$duration_min, 2)`-`r round(durations$spa_eng$duration_max, 2)`).

@@

For each translation pair, we defined three predictors of interest: the lexical frequency of the correct translation (*Frequency*), the phonological similarity between the presented (Catalan or Spanish) word and their correct English translation (*Similarity*), and the presented word's number of cross-language phonological neighbours (*CLPN*).

We included *Frequency* as a nuance predictor, under the hypothesis that---keeping other predictors constant---participants would translate higher-frequency words more accurately and faster than lower-frequency words. Lexical frequencies of correct translations were extracted from SUBTLEX-UK [@van2014subtlex], and transformed to Zipf scores. Words in the stimuli list without a lexical frequency value were excluded from data analysis (`r stim_exc_freq$cat_eng$n` in the Catalan list, `r stim_exc_freq$spa_eng$n` in the Spanish list).

We calculated *Similarity*, our main predictor of interest, by computing the Levenshtein similarity between the X-SAMPA transcriptions of each pair of translations using the `stringdist` R package [@van2014stringdist]. The Levenshtein distance computes the edit distance between two character strings---in this case, phoneme transcriptions---by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. We divided this edit distance by the number of characters included in the longest X-SAMPA transcription of the translation pair. This results in a proportion score, in which values closer to zero correspond to lower Levenshtein distances between phonological transcriptions (i.e., higher similarity), and values closer to 1 correspond to higher Levenshtein distances (i.e., lower similarity). This transformation also help account for the fact that the Levenshtein distance tends to increase with the length of the transcriptions. For interpretability, we subtracted this proportion from one, so that values closer to one correspond to higher similarity between phonological transcriptions, and values closer to zero correspond to lower similarity between phonological transcriptions. For example, the *table* (`teIb@l`)-*mesa* (`mesa`) translation pair had a `r scales::percent(stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (`trEIn`)-*tren* (`tREn`) translation pair had a `r scales::percent(stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity. @tbl-stimuli summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

@@

```{r tbl-stimuli}
#| label: tbl-stimuli
#| tbl-cap: "Stimuli properties."
stimuli |> 
    summarise(across(c(freq_zipf_2, lv, neigh_n_h),
                     lst(mean, sd, min, max), 
                     na.rm = TRUE),
              .by = c(group)) |> 
    gt(rowname_col = "group") |> 
    tab_spanner("Frequency", matches("freq")) |> 
    tab_spanner("Similarity", matches("lv")) |> 
    tab_spanner("CLPN", matches("neigh")) |> 
    fmt_number(is.numeric) |> 
    fmt_integer(c(neigh_n_h_min,
                  neigh_n_h_max)) |> 
    cols_merge_uncert(freq_zipf_2_mean, freq_zipf_2_sd) |> 
    cols_merge_uncert(lv_mean, lv_sd) |> 
    cols_merge_uncert(neigh_n_h_mean, neigh_n_h_sd) |> 
    cols_merge_range(freq_zipf_2_min, freq_zipf_2_max) |> 
    cols_merge_range(lv_min, lv_max) |> 
    cols_merge_range(neigh_n_h_min, neigh_n_h_max) |> 
    cols_label(freq_zipf_2_mean = "Mean ± SD",
               freq_zipf_2_sd = "SD",
               freq_zipf_2_min = "Range",
               lv_mean = "Mean ± SD",
               lv_sd = "SD",
               lv_min = "Range",
               neigh_n_h_mean = "Mean ± SD",
               neigh_n_h_sd = "SD",
               neigh_n_h_min = "Range") |> 
    tab_style(cell_text(weight = "bold"),
              cells_column_spanners()) |> 
    tab_style(cell_text(style = "italic"),
              cells_column_labels()) |> 
    grand_summary_rows(columns = matches("mean|sd"),
                       fns = list(label = md("**Mean**"), 
                                  id = "totals",
                                  fn = "mean"),
                       fmt = ~fmt_number(., n_sigfig = 3, suffixing = TRUE))
```

@@

![Schematic representation of a trial in the experimental task](_assets/img/design.png)

@@

```{r dataset-1}
#| label: dataset-1
dataset <- exp_responses |> 
    filter(group!="cat-SPA") |> 
    group_split(group) |> 
    set_names(c("spa-ENG", "cat-ENG")) |> 
    janitor::clean_names()

>>>>

# import objects
tar_config_set(
  store = here::here("_targets"),
  script = here::here("_targets.R")
)
tar_load_globals()
tar_load(participants)
tar_load(exp_participants)
tar_load(exp_responses)
tar_load(quest_responses)
tar_load(quest_participants)
tar_load(stimuli)
tar_load(epreds)
tar_load(dataset_1)
tar_load(dataset_2)
tar_load(dataset_3)
tar_load(exp_1_m0)
tar_load(exp_2_m0)
tar_load(exp_3_m1)

theme_set(theme_ggdist() +
  theme(panel.grid.major.y = element_line(
    colour = "grey",
    linetype = "dotted"
  )))

@@

When some German speakers listened to the song *The Power* by SNAP! (*Coyote Ugly*, 2000), many of them misheard the line "I've got the power" as *Agatha Bauer!* When Michael Jackson vocalised "ma-ma-coo-sah" in his song *Wanna Be Startin' Somethin'* (*Thriller*, 1982), some Dutch speakers understood *mama appelsap* [mama applejuice]. When Spanish speakers listen to the line "Circumvent your thick ego" from the song *Pictures* by System of a Down (*Steal this Album!*, 2002), they tend to hear *Sácame de aquí, King-Kong* [take me out of here, King-Kong]. When Japanese speakers listen to the line "I want to hold your hand" in the homonym song by The Beatles (*Meet The Beatles*, 1964), they often mishear *Aho na hōnyōhan* [stupid urinator]. Outrageous as these examples might sound to speakers of other languages, many readers may know a few cases in their own native language. This auditory illusion can feel quite real, and often inevitable after the first time it happens. In Japanese, this phenomenon takes the name *Soramimi* (lit. "empty ear"), after the *Soramimi Hour*---a section of the program *Tamori Club* hosted by comedian Morita Kazuyoshi (Tamori)---in which instances of this illusion were presented with comedy purposes. Soramimi is a particular case of *homophonic translation*: words or phrases in one language are translated into similar-sounding words and phrases in a different language, without necessarily preserving the meaning. This phenomenon extends beyond song lyrics, and may occur anytime when listening to non-native speech. For instance, when listening to "You know me?" (\textipa{/ju: n@U mi:}/) in English, Japanese-native listeners tend to understand *yunomi* (/\textipa{jUnomi}/), which is a type of Japanese teacup. Overall, the occurrence of Soramimi---and more generally, homophonic translations---suggests that speech in an unfamiliar language has the potential to activate lexical representations in the native language.

Although homophonic translations do not necessarily take place with the preservation of meaning---which often results in outrageous mistranslations like the ones aforementioned---homophonic translations sometimes may lead to correct translations. For example, imagine a Spanish native with no previous familiarity with any other language listening to French for the first time. This listener may encounter the word /\textipa{"pOKt}/ (*porte*), translation of door in French. When comparing the phonemic transcription of the French to that of its Spanish translation /\textipa{"pweR.ta}/ (*puerta*), one may be led to infer that these two words do not have much overlap. But one thing to note is that the voiced uvular fricative consonant /\textipa{K}/ and the open-mid back rounded vowel /\textipa{O}/, which do not exist in Spanish as phonemic categories, can be perceived as allophonic variations the native (Spanish) phonemes /\textipa{R}/ and /o/, as described by assimilation models of phoneme perception [@best2001discrimination; @best1988examination]. Although this phonetic mismatch between the non-native word and its translation can have a noticeable toll on comprehension @cutler2004patterns, the Spanish listener may still be able to activate the phonological representation of /\textipa{"pweR.ta}/ (*puerta*) from hearing the word-form /\textipa{"pOKt}/ (*porte*), ultimately allowing them to access the correct semantic representation. Given enough phonological similarity between the non-native word and its translation in the native language, a naïve listener may succeed at word comprehension when listening to an unfamiliar language. Such phonological similarity between translation equivalents---known as *cognateness*---is common across many languages, and is often due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents  [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and orthographic similarity of the former to the equivalent in their native language.

While Soramimi (and more generally homophonic translations) are popular and frequent phenomena, the psycholinguistic literature on the topic is relatively scarce, and has mostly focused on the phonological variables involved in the occurence of Soramimi. For instance, @otake2007interlingual analysed 194 instances of Soramimi, broadcasted between 1992 and 2007 by the TV show Soramimi hour. These instances consisted of homophonic translations of English song lyrics to Japanese words and phrases. The vast majority of the reported instances were phrases (96%) and a few, to single words (4%). The phonetic features of the presented English lyrics were preserved with relatively high variability, with some Japanese resulting words or phrases sharing high overlap with the original English strings of sounds, and some sharing very little overlap. When analysing the few instances of homophonic translations of single words, the author identified three phonological processes that might explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /\textipa{kraI}/ to *kurai*, /\textipa{kuRai}/ [dark]), deletion (e.g., *go*, /\textipa{goU}/ to *go*, /go/ [go], and alternation (e.g., *low*, /\textipa{loU}/ to *rou*, /\textipa{Roo}/ [wax]). This suggests that the reported homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to the Japanese phoneme inventory and Japanese phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic]. Native prosody also constrains the occurence of homophonic translations. For instance, @kentner2015rhythmic presented in the auditory modality French-native and German-native participants with English songs with lyrics. Homophonic translations followed language-specific segmentation strategies. Most homophonic translations reported by German-native participants resulted from inserting phrase and word boundaries *before* stressed syllables (following the more frequent trochaic patter in German). On the contrary, most homophonic translations reported by French-native participants resulted from inserting phrase and word boundaries *after* stressed syllables, following the more frequent iambic pattern in French.

These studies have addressed the phonological aspects of homophonic translation, but little attention has been paid to the dynamics of lexical activation and competition that underlay this phenomenon. In the present work, we investigated the interplay between phonological similarity and phonological neighbourhood density in a non-native speech perception task. In particular, we studied how the presence of phonological neighbours in the native language lexicon affects the extent to which listeners are able to activate the correct translation of an auditorily presented word in thier native language. We used a translation elicitation task in which participants listened to words from an unfamiliar language, and then tried to guess their translation in their native language. We will henceforth refer to the auditory-presented words heard by participants on each trial as *presented words*, and the correct translation for the presented words as *target translations*. We explored listeners’ reliance on phonological similarity to succeed in the task by manipulating the amount of phonological similarity between the presented words and their target translations. Since participants were unfamiliar with the presented language, they could only use phonological similarity between the presented language and their native language to successfully translate the words. We therefore predicted that participants’ performance should increase when the translation pairs are phonologically more similar. We also predicted that there is a minimum threshold of phonological similarity to be sufficient for translation.

@@

In Experiment 1, we collected data from two groups of British English natives living in the United Kingdom. One group was presented with Catalan words, the other with Spanish words. We examined the extent to which participants were able to use the phonological similarity between the presented word (in Catalan or Spanish) and its target translation to provide accurate responses. In Experiment 2, we tested a group of (European) Spanish natives in the same task, who were presented with Catalan words. Catalan and Spanish are two Romance languages whose close typological distance is reflected in the fact that they share many cognates, where English is a Germanic language that shares considerably fewer cognates with Catalan and Spanish. By testing participants translating words from typologically close or distant languages, we expected to widen the range of the phonological similarity scores of the translation pairs involved in the experimental task, therefore allowing us to explore potential cross-language differences in participant’s performance. One unexpected finding was that participants in Experiments 1 and 2 were surprisingly good at translating a subset of words which had low phonological similarity with their correct translation. We were concerned that this may be caused by some prior knowledge of specific words by our participants, as some Spanish words are commonly seen in media or product labels, making them familiar even to monolingual speakers of English. In Experiment 3, we collected additional data from a new group of British English natives. The design was closely modelled after Experiment 1, except that after providing their response in each trial, participants reported whether they had previous knowledge of the presented word. In the final section of this article, we present analyses on the joint data sets of Experiments 1 to 3.

@@

```{r}
#| label: participants-numbers-1
participants_1 <- exp_participants |>
  filter(group != "cat-SPA")

@@

n_participants <- n_distinct(participants_1)
n_participants_group <- table(participants_1$group)
participants_age <- participants_1 |>
  summarise(across(age, lst(mean, sd, min, max)))
n_female <- table(participants_1$gender)
```
We collected data from `r n_participants` British English-native adults living in United Kingdom (*Mean* = `r round(participants_age$age_mean, 2)` years, *SD* = `r round(participants_age$age_sd, 2)`, Range = `r round(participants_age$age_min, 2)`-`r round(participants_age$age_max, 2)`, `r n_female[1]` female). Data collection took place from `r collection_dates[1]` to `r collection_dates[2]`. Participants were recruited via Prolific (5£ compensation) and SONA (compensation in academic credits). Participants gave informed consent before providing any data. The study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was by the University of Oxford Medical Sciences Inter-Divisional Research Ethics Committee (IDREC) (R60939/RE007). Participants were asked to complete the experiment using a laptop in a quiet place with good internet connection.

```{r tbl-participants}
#| label: tbl-participants
#| tbl-cap: "Participant details."
tbl_data <- participants |>
  mutate(exp = case_when(
    source == "Experiment" & group != "cat-SPA" ~ "Experiment 1",
    source == "Experiment" & group == "cat-SPA" ~ "Experiment 2",
    source == "Questionnaire" ~ "Experiment 3",
  )) |>
  summarise(
    n = n(),
    n_excluded = sum(!valid_status == "Valid"),
    across(age, lst(mean, sd, min, max)),
    l2_lst = list(l2),
    .by = c(group, exp)
  ) |>
  mutate(
    l2_lst = map(
      l2_lst,
      function(x) {
        y <- table(x)
        y <- y[names(y) != "None"]
        y <- paste0(names(y), " (", y, ")")
        return(y)
      }
    ),
    n_excluded = paste0(n, " (", n_excluded, ")")
  ) |>
  select(-n)

tbl_data |>
  gt(groupname_col = "exp", rowname_col = "group") |>
  fmt_number(matches("age")) |>
  fmt_integer(matches("min|max")) |>
  cols_merge_uncert(age_mean, age_sd) |>
  cols_merge_range(age_min, age_max) |>
  tab_spanner("Age", matches("age")) |>
  cols_label(
    group = "Group",
    exp = "Experiment",
    n_excluded = "N",
    age_mean = "Mean ± SD ",
    age_min = "Range",
    l2_lst = "L2"
  ) |>
  tab_style(
    cell_text(style = "italic"),
    cells_column_labels()
  ) |>
  tab_style(
    cell_text(weight = "bold"),
    cells_column_spanners()
  ) |>
  tab_style(
    cell_text(align = "left"),
    list(cells_body(), cells_column_labels())
  ) |>
  tab_style(
    cell_text(
      weight = "bold",
      align = "center",
      style = "normal"
    ),
    cells_column_labels(c(l2_lst, n_excluded))
  ) |>
  tab_footnote(
    "Number of included participants (number of excluded participants.)",
    locations = cells_column_labels(n_excluded)
  )
```

@@

```{r}
#| label: stimuli-lengths
lengths <- stimuli |>
  select(group, ipa_flat_1, ipa_flat_2, word_1, word_2) |>
  mutate(across(ipa_flat_1:word_2, nchar)) |>
  summarise(
    across(
      c(ipa_flat_1, ipa_flat_2, word_1, word_2),
      lst(mean, sd, min, max)
    ),
    .by = group
  )
lengths <- split(lengths, lengths$group)
names(lengths) <- janitor::make_clean_names(names(lengths))

durations <- stimuli |>
  summarise(across(duration, lst(mean, sd, min, max)),
    .by = group
  )
durations <- split(durations, durations$group)
names(durations) <- janitor::make_clean_names(names(durations))
```

We created two lists of input words to be presented to participants in the auditory modality: one in Catalan and one in Spanish. Words in the Catalan list were `r round(lengths$cat_eng$ipa_flat_1_mean, 2)` phonemes long on average (*SD* = `r round(lengths$cat_eng$ipa_flat_1_sd, 2)`, Range = `r round(lengths$cat_eng$ipa_flat_1_min, 2)`-`r round(lengths$cat_eng$ipa_flat_1_max, 2)`), and the orthographic forms of their English translations (which participants had to type) were `r round(lengths$cat_eng$word_2_mean, 2)` characters long on average (*SD* = `r round(lengths$cat_eng$word_2_sd, 2)`, Range = `r round(lengths$cat_eng$word_2_min, 2)`-`r round(lengths$cat_eng$word_2_max, 2)`). Words in the Spanish list were `r round(lengths$spa_eng$ipa_flat_1_mean, 2)` phonemes long on average (*SD* = `r round(lengths$spa_eng$ipa_flat_1_sd, 2)`, Range = `r round(lengths$spa_eng$ipa_flat_1_min, 2)`-`r round(lengths$spa_eng$ipa_flat_1_max, 2)`), and the orthographic form of their English translations were `r round(lengths$spa_eng$word_2_mean, 2)` characters long on average (*SD* = `r round(lengths$spa_eng$word_2_sd, 2)`, Range = `r round(lengths$spa_eng$word_2_min, 2)`-`r round(lengths$spa_eng$word_2_max, 2)`).

In each trial, participants listened to one audio file, wich contained a single word. The audio files were the same ones used in child experiments conducted in the Laboratori de Recerca en Infància of Universitat Pompeu Fabra (Barcelona, Spain). These audio files were recorded by a proficient Catalan-Spanish female bilingual from the Metropolitan Area of Barcelona in a child-directed manner. Catalan and Spanish words were recorded at 44,100 Hz in separate files in the same session, and then de-noised using Audacity and normalised at peak intensity using Praat [@broersma2021praat]. The average duration of the Catalan audio files was `r round(durations$cat_eng$duration_mean, 2)` seconds (*SD* = `r round(durations$cat_eng$duration_sd, 2)`, Range = `r round(durations$cat_eng$duration_min, 2)`-`r round(durations$cat_eng$duration_max, 2)`). The average duration of the Spanish audio files was `r round(durations$spa_eng$duration_mean, 2)` seconds (*SD* = `r round(durations$spa_eng$duration_sd, 2)`, Range = `r round(durations$spa_eng$duration_min, 2)`-`r round(durations$spa_eng$duration_max, 2)`).

@@

For each word in the Catalan and Spanish lists pair, we defined three predictors of interest: the lexical frequency of the correct English translation (*Frequency*), the phonological similarity between the presented word (in Catalan or Spanish) and their correct English translation (*Similarity*), and the presented word's number of cross-language phonological neighbours (*CLPN*) in English. *Frequency* was included as a nuance predictor, under the hypothesis that---keeping other predictors constant---participants would translate higher-frequency words more accurately and faster than lower-frequency words. Lexical frequencies of correct translations were extracted from SUBTLEX-UK [@van2014subtlex], and transformed to Zipf scores. Words in the stimuli list without a lexical frequency value were excluded from data analysis (`r stim_exc_freq$cat_eng$n` in the Catalan list, `r stim_exc_freq$spa_eng$n` in the Spanish list).

We calculated *Similarity*, our main predictor of interest, by computing the Levenshtein similarity between the X-SAMPA transcriptions of each pair of translations using the `stringdist` R package [@van2014stringdist]. The Levenshtein distance computes the edit distance between two character strings---in this case, two phonological transcriptions---by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. We divided this edit distance by the number of characters included in the longest X-SAMPA transcription of the translation pair. This results in a proportion score, in which values closer to zero correspond to lower Levenshtein distances between phonological transcriptions (i.e., higher similarity), and values closer to 1 correspond to higher Levenshtein distances (i.e., lower similarity). This transformation accounts for the fact that the Levenshtein distance tends to increase with the length of the transcriptions. For interpretability, we subtracted this proportion from one, so that values closer to one correspond to higher similarity between phonological transcriptions, and values closer to zero correspond to lower similarity between phonological transcriptions. For example, the *table* (`teIb@l`)-*mesa* (`mesa`) translation pair had a `r scales::percent(stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (`trEIn`)-*tren* (`tREn`) translation pair had a `r scales::percent(stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity. @tbl-stimuli summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

@@

```{r tbl-stimuli}
#| label: tbl-stimuli
#| tbl-cap: "Stimuli details."
stimuli |>
  summarise(
    across(c(freq_zipf_2, lv, neigh_n_h),
      lst(mean, sd, min, max),
      na.rm = TRUE
    ),
    .by = c(group)
  ) |>
  gt(rowname_col = "group") |>
  tab_spanner("Frequency", matches("freq")) |>
  tab_spanner("Similarity", matches("lv")) |>
  tab_spanner("CLPN", matches("neigh")) |>
  fmt_number(is.numeric) |>
  fmt_integer(c(
    neigh_n_h_min,
    neigh_n_h_max
  )) |>
  cols_merge_uncert(freq_zipf_2_mean, freq_zipf_2_sd) |>
  cols_merge_uncert(lv_mean, lv_sd) |>
  cols_merge_uncert(neigh_n_h_mean, neigh_n_h_sd) |>
  cols_merge_range(freq_zipf_2_min, freq_zipf_2_max) |>
  cols_merge_range(lv_min, lv_max) |>
  cols_merge_range(neigh_n_h_min, neigh_n_h_max) |>
  cols_label(
    freq_zipf_2_mean = "Mean ± SD",
    freq_zipf_2_sd = "SD",
    freq_zipf_2_min = "Range",
    lv_mean = "Mean ± SD",
    lv_sd = "SD",
    lv_min = "Range",
    neigh_n_h_mean = "Mean ± SD",
    neigh_n_h_sd = "SD",
    neigh_n_h_min = "Range"
  ) |>
  tab_style(
    cell_text(weight = "bold"),
    cells_column_spanners()
  ) |>
  tab_style(
    cell_text(style = "italic"),
    cells_column_labels()
  ) |>
  grand_summary_rows(
    columns = matches("mean|sd"),
    fns = list(
      label = md("**Mean**"),
      id = "totals",
      fn = "mean"
    ),
    fmt = ~ fmt_number(., n_sigfig = 3, suffixing = TRUE)
  )
```

@@

![Schematic representation of a trial in the experimental task. The trial stars with the presentation of a fixation point in the centre of the screen (yellow dot). After 1,000 ms, the auditory word was presented while the fixation point remained on screen. After the offset of the audio, the fixation point disappeared and a visual prompt (`>`) was presented. Participants then wrote their and clicked `RETURN`, making the end of the trial.](_assets/img/design.png)

@@

```{r dataset-1}
#| label: dataset-1
dataset <- exp_responses |>
  filter(group != "cat-SPA") |>
  group_split(group) |>
  set_names(c("spa-ENG", "cat-ENG")) |>
  janitor::clean_names()
